---
title: "Chapter_3_Notes_Reproducibles"
format: html
editor: source
---

## Reading Help Files

Can use function:

?function_name help(function_name)

to get help on a specific function you don't understand

> ?write.table

will load in a help page in Rstudio

The sections the work page is broken down into are:

**Description**: An extended description of what the function does.

**Usage**: The arguments of the function and their default values (which can be changed).

**Arguments**: An explanation of the data each argument is expecting.

**Details**: Any important details to be aware of.

**Value**: The data the function returns.

**See Also**: Any related functions you might find useful. \* Examples: Some examples for how to use the function.

some have same help file \> ?write.table() ?write.csv()

## Special Operators

Use: ?"\<-" ?`<-`

## Getting help with packages

Many packages include "vignettes," which are tutorials and extended example documentation.

To list all vignettes for installed packages, use vignette() without any arguments.

To list all available vignettes for a specific package, use vignette(package="package-name").

To open a specified vignette, use vignette("vignette-name").

If a package doesn’t have any vignettes, you can typically find help by typing help("package-name").

RStudio offers a set of excellent cheatsheets for many packages.4

## When you remember part of the funtion name

You can perform a fuzzy search in R to find functions even if you're unsure about the exact package or spelling.

To conduct a fuzzy search, use ??function_name in R.

Fuzzy search involves searching for an approximate string match.

For instance, if you recall that a function related to setting the working directory contains "set" in its name but aren't sure of the exact function name, you can use a fuzzy search to identify it.

Example: To find functions related to setting the working directory, you can use **??set in R**

## Randy Swaty Notes on chapter 1-3:

PUll when first opening a callabe with other people

## Most common things to look for when you get errors

1.  Reading in data

```{r}
##data <- read.csv('dataset.csv')
```

> Solution: keep everything in a r-studio project and check for typos google for the dataset check directory use TAB for autofill to help prevent typos

2.  Packages

```{r}
##library(tidyverse)

##df <- cars %>%
  ##filter
```

-   always check for packages
-   google the function or obv the error
-   remember you can define which packages to get function for using

commute stage commit then push

chapter: exploring data frames

srt(gapminder)

-   shows the structure of the data frame

head \* will show the first 6 rows of the data set

Tail \* looks at the last 6 rows of the data

google: base vs tidyverse functions
